Hydra Battles in Coq

### 2020-10-28 doc/hydras.tex
typos in the pdf file

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, Bordeaux-INP\\with contributions by \'Evelyne Contejean,  Florian Hatat and Pascal Manoury}
\date{\today}
\title{Hydra battles in Coq}
\makeindex
\begin{document}
\maketitle

@@

\subsection{Difference from Kirby and Paris's work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

\subsection{Trust in our proofs?}
\label{sect:trust-in-proofs}

@@

\subsection{Assumed redundancy}

@@

\subsection{Main references}

@@

\subsection{State of the development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on  notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

@@

\subsection{Typographical conventions}

@@

\subsection{Active links}
The  links which appear in this pdf  document lead are of three possible kinds of destination:
\begin{itemize}
\item Local links to the document itself,
\item External links, mainly to \coq's page,
\item Local links to pages generated by \texttt{coqdoc}. According to the current makefile,
  we assume that the page generated from a library \texttt{XXX/YYY.v} is stored as
the relative address \texttt{../src/html/hydras.XXX.YYY.html} (from the location of the pdf).
Thus,  active links towards our \coq{} modules may be incorrect if you got this \texttt{pdf} document otherwise than by compiling the distribution available in
\url{https://framagit.org/casteran/hydra-battles-in-coq}.

@@

\subsection{Alternative or bad definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

@@

\subsection{The rules of the game}

@@

\section{Hydras and their representation in \emph{Coq}}
\label{sec:orgheadline48}

@@

\subsubsection{Recursive functions on type \texttt{Hydra}}
\label{sec:orgheadline41}
\label{sec:hsize-def}

@@

\subsection{Induction principles for hydras}
\label{sec:orgheadline42}

@@

\subsubsection{A failed attempt}

@@

\subsubsection{A principle for  mutual induction}
In order to get an appropriate induction scheme for the types 
\texttt{Hydra} and \texttt{Hydrae}, we can use  \coq{}'s  command \texttt{Scheme}.

@@

\subsubsection{A correct proof}

@@

\section{Relational description of hydra battles}

@@

\subsection{Chopping off a head at distance 1 from the foot (relation  R1)}

@@

\subsection{Chopping off a head at distance \texorpdfstring{$\geq 2$}{>= 2} from the foot (relation R2) }

@@

\subsection{Relation associated with a round}

@@

\subsection{Rounds and battles}

@@

\subsection{Classes of battles}
\label{sect:battle-classes}

@@

\subsection{Big steps}

@@

\section{A long battle}
\label{sect:big-battle}

@@

\subsection{The beginning of hostilities}
During the two first rounds, our hydra loses its two rightmost heads. Thus just before the third round, it looks like in figure~\vref{fig:hinit-plus2}.

@@

\subsection{Looking for regularities}

@@

\subsection{Giant steps}

@@

\subsection{A minoration lemma}

@@

\section{Generic properties}

@@

\subsection{A  small proof of impossibility}
\index{Maths!Proofs of impossibility}

@@

\chapter{Introduction to ordinal numbers and ordinal notations}

@@

\section{The mathematical point of view}

\subsection{Well-ordered sets}
Let us start with some definitions.
A  \emph{well-ordered set} is a set provided with a binary relation $<$ which has the following properties.
\begin{description}
\item[irreflexivity] : $\forall x\in A, x\not< x$
\item[transitivity] : $\forall x\,y\,z\in A, x<y \Rightarrow y<z \Rightarrow x<z$
\item[trichotomy]: $\forall x\,y\in A, x<y \vee x = y \vee y < x$
\item[well foundedness]: $<$ is well-founded (every element of $A$ is accessible)\footnote{In classical mathematics, we would say that there is no infinite sequence $a_1>a_2> \dots a_n> a_{n+1}\dots$ in $A$. Please refer to any documenation on \coq{} for having more details on well-foundedness and accessibility.}.

@@

\subsection{Ordinal numbers}

@@

\section{Ordinal numbers in Coq}

@@

\section{Countable ordinals}

@@

\section{Ordinal notations}

@@

\subsubsection*{Ordered types}

@@

\subsection{A class for ordinal notations}

@@

\subsection{Ordinal notations and measures for proving termination}
\label{sect:measure-ON}

@@

\section{Examples of ordinal notations}

\subsection{The ordinal \texorpdfstring{$\omega$}{omega}}

@@

\subsection{Sum of two ordinal notations}

@@

\subsection{The ordinal \texorpdfstring{$\omega+\omega$}{omega + omega}}

@@

\subsection{Limits and successors in \texorpdfstring{$\omega+\omega$}{omega+omega}}

@@

\section{The ordinal \texorpdfstring{$\omega^2$}{omega^2}}

@@

\subsection{A proof of termination using \texorpdfstring{$\omega^2$}{omega^2}} 
\label{sect:merge-example}.

@@

\subsection{Yet another  proof of impossibility}
\label{omega2-case}

@@

\subsubsection{Proof of the inequality \texttt{m small\_h o< m big\_h}}

@@

\subsubsection{Proof of the inequality \texttt{m big\_h o<= m small\_h} }

@@

\subsubsection{End of the proof}

@@

\section{A notation for finite ordinals}

@@

\section{Comparing two ordinal notations}

@@

\section{Comparing an ordinal notation with Schütte's model}

@@

\section{Isomorphism of ordinal notations}

@@

\section{Other ordinal notations}

@@

\chapter[A proof of termination, using epsilon0]{A proof of termination, using ordinals below \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\section{The ordinal \texorpdfstring{\(\epsilon_0\)}{epsilon0}}
\label{sec:epsilon0-intro}

\subsection{Cantor normal form}
\index{Maths!Cantor normal form}

@@

\subsection{A data type for  ordinals in Cantor normal form}
\label{sec:orgheadline72}
\label{sec:T1-inductive-def}

@@

\subsubsection{Finite ordinals}
\label{sec:orgheadline67}

@@

\subsubsection{The ordinal \(\omega\)}
\label{sec:orgheadline68}

@@

\subsubsection{The ordinal \(\omega^\alpha\), a.k.a. \(\phi_0(\alpha)\)}
\label{sec:orgheadline70}

@@

\subsubsection{The hierarchy of \(\omega\)-towers:}
\label{sec:orgheadline71}

@@

\subsection{Comparison between ordinal terms}
\label{sec:orgheadline73}

@@

\subsubsection{A predicate for characterizing normal forms}
\label{sect:t1-nf}

@@

\subsection{Making normality implicit}
  We would like to get rid of terms of type \texttt{T1} which are not in Cantor normal form.
A simple way to do this is to consider statements of the form 
\texttt{forall alpha: T1, nf alpha -> $P$ alpha}, where $P$ is a predicate over type \texttt{T1}, like in the following lemma \footnote{Ordinal addition is formally defined a little later (page~\ref{sect:infix-plus-T1})}.

@@

\subsubsection{A sigma type for \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\subsection{Syntactic definition of limit and successor ordinals}

@@

\subsubsection{Addition and multiplication}

@@

\subsection{Pretty printing Ordinals in Cantor normal Form}
\label{sect:ppT1}

@@

The following data type defines a abstract syntax for more readable ordinals in Cantor normal form:

@@

\subsubsection{Arithmetic on type \texttt{E0}}

@@

\section{Well-foundedness and transfinite induction}

@@

\subsection{About  well-foundedness}
\label{sec:orgheadline82}
   In order to use \texttt{T1} for proving termination results,
we need to prove that  our order \texttt{<} is well-founded. Then we will get \emph{transfinite induction} for free.

@@

\subsubsection{A first attempt}
\label{sec:orgheadline77}
\index{Coq!Techniques!Well-founded induction}

@@

\subsubsection{Using a stronger inductive predicate.}
\label{sec:orgheadline78}
  Instead of trying to prove directly that any ordinal term \(\alpha\) in normal form is accessible
through \texttt{LT}, we propose to show first that any well formed 
term of the form \(\omega^\alpha\times(n+1)+\beta\) is accessible (which is a stronger result).

@@

\subsection{An ordinal notation for  \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\section{A variant for hydra battles}

\subsection{Natural sum (a.k.a. Hessenberg's  sum)}
\label{sec:orgheadline87}
\label{hydra-variant}

@@

\subsubsection{Rewriting lemmas}
\label{sec:orgheadline86}

@@

\index{Projects}
\begin{project}
Compare \texttt{oplus}'s definition (with inner fixpoint) with other possibilities
(\texttt{coq-equations}, \texttt{Function}, etc.).
\end{project}
\subsection{More theorems on Hessenberg's sum}

@@

\subsubsection{Commutativity, associativity}

@@

\subsection{A measure for hydra-battle termination}

@@

\chapter[The Ketonen-Solovay machinery]{Strolling inside \texorpdfstring{$\epsilon_0$}{epsilon0}: The Ketonen-Solovay machinery\label{ks-chapter}}
\label{chap:ketonen}
\index{Maths!Ketonen-Solovay machinery}

@@

\subsubsection{Mathematical definition of canonical sequences}

@@

\subsubsection{Canonical sequences in Coq}

@@

\subsection{Basic properties of canonical sequences}

@@

\subsubsection{Canonical sequences and the order $<$}

@@

\subsubsection{Limit ordinals are really limits}
The following theorem states that any limit ordinal $\lambda<\epsilon_0$ 
is the limit of the sequence \showmath{\canonseq{\lambda}{i}\;(1\le i)}.

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : paths}
\index{Maths!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\subsection{Formal definition}

@@

\subsection{Existence of a path}

@@

\subsection{Paths and hydra battles}
\label{KS-o2h}

@@

\section{A  proof of impossibility}

@@

\subsection{The case of free battles}
\label{sec:free-battles-case}
Let us now consider that $B$ is instantiated to \texttt{free} (which means that we are considering proofs of termination of \emph{all} battles). The following lemmas are proved in Module~\href{../src/html/hydras.Hydra.Epsilon0_Needed_Free.html}{Hydra.Epsilon0\_Needed\_Free}.
The case $B=\texttt{standard}$ is studied in section~\vref{std-case}.

@@

\section{The case of standard battles}
\label{sec:standard-intro}\label{std-case}
One may wonder if our theorem holds also in the framework of standard battles. Unfortunately, its proof relies on the lemma \texttt{LT\_to\_round\_plus} of
Module~\href{../src/html/hydras.Hydra.O2H.html}{Hydra.O2H}.

@@

\subsection{Paths with constant index}

@@

\subsubsection{Sketch of proof of \texttt{Cor12}}
\index{Maths!Transfinite induction}

@@

\subsection{Casting paths with constant index into standard paths}

@@

\subsubsection{Sketch of proof of \texttt{constant\_to\_standard\_path}}

@@

\subsection{Back to hydras}
\label{sec:standard-battles-cases}
We are now able to complete our proof that there exists no bounded variant for proving the termination of standard hydra battles. This proof can
be consulted in the module 
\url{../src/html/hydras.Hydra.Epsilon0_Needed_Std.html}.
Please note that it has the same global structure as in section\ref{sec:free-battles-case} 
% ICI !
Applying the  lemmas  \texttt{Lemma2\_6\_1} of the module 
\href{../src/html/hydras.Epsilon0.Paths.html\#Lemma2_6_1}%
{\texttt{Lemma2\_6\_1}}   and 
\href{../src/html/hydras.Epsilon0.Paths.html\#constant_to_standard_path}%
{\texttt{constant\_to\_standard\_path}},
we can convert any inequality $\beta<\alpha<\epsilon_0$ into a standard path from
$\alpha$ to  $\beta$, then into a fragment of a standard battle from 
$\iota(\alpha)$ to $\iota(\beta)$.

@@

%---------------------------------------------------------------------
\chapter{Large sets and rapidly growing functions}\label{chap:alpha-large}

@@

\section{The length of minimal large sequences}

@@

\subsection{Formal specification}

@@

\subsection{Abstract properties}

@@

\subsubsection{The  ordinal zero}
\label{sect:L-spec-zero}
The base case is directly a consequence of the specification.

@@

\subsubsection{Successor ordinals}
\label{sect:L-spec-succ}
Let $\beta$ be some ordinal, and assume the arithmetic function $f$ satisfies 
the specification $(\texttt{L\_spec}\;\beta)$.  Let $k$ be any natural number.
Any path from $\texttt{succ}\,\beta$ to $0$ starting at $k+1$ can be decomposed into a first step from $\texttt{succ}\,\beta$ to $\beta$, then a path from
$\beta$ at $k+2$ to $0$. 
By hypothesis the interval $[k+2, f(k+2)-1]$ is $\beta$-mlarge.
But the interval $[k+1, f(k+2)-1]$ is the concatenation of the singleton
$\{k+1\}$ and the interval $[k+2, f(k+2)-1]$.
So, the function $\lambda\,k.\,f(k+1)$ satisfies the specification $\texttt{L\_spec}\,\beta$.

@@

\subsubsection{Limit ordinals}
\label{sect:L-spec-lim}

@@

\subsection{First results}

@@

\subsubsection{Finite ordinals}

@@

\subsubsection{The first limit ordinal \texorpdfstring{$\omega$}{omega}}

@@

\subsubsection{Going further}
Let us consider a last example, ``computing'' $L_{\omega^3}$.
Since the canonical sequence associated with this ordinal is composed of the
$\omega^2\times i\;(i\in\mathbb{N}_1)$, we have to study this sequence.

@@

\subsubsection{Back to Hydra battles}

@@

\section{The Wainer-Hardy hierarchy (functions \texorpdfstring{$H_\alpha$}{\texttt{H\_alpha}})}

@@

\subsection{Hardy functions in \texttt{Coq}}

@@

\subsection{First  steps of the Hardy hierarchy}
Using rewrite rules from \texttt{H\_eq1} to \texttt{H\_eq4}, we can explore the functions $H_\alpha$ for some small values of $\alpha$.

\subsubsection{Finite ordinals}

@@

\subsubsection{New limits}

@@

\subsubsection{A numerical example}

@@

\subsection{Abstract properties of H-functions}
~\label{sect:H-alpha-prop}

@@

\subsubsection{Back to hydras}

@@

\section{The Wainer hierarchy (functions \texorpdfstring{$F_\alpha$}{F\_alpha})}
\label{sect:wainer}

@@

%----------------------------------------------------------------------
\chapter[Countable Ordinals (after Sch\"{u}tte)]{Kurt Schütte's axiomatic Definition of countable Ordinals}

@@

\section{Additional  axioms}

@@

\subsubsection{Classical logic}

@@

\subsubsection{Description operators}

@@

\subsubsection{The definition of zero}

@@

\section{The  successor function}

@@

\section{Finite ordinals}

@@

\section{The definition of \texttt{omega}}
In order to define $\omega$, the first infinite ordinal, we use an operator which
``returns'' the least upper bound (if it exists) of a subset $X\subseteq \mathbb{O}$.
For that purpose, we first use a predicate:
(\texttt{is\_lub $D$ \textit{lt} $X$ $a$}) if $a$ belongs to $D$ and is the least 
upper bound  of $X$ (with respect to \textit{lt}).

@@

\subsection{Ordering functions and ordinal addition}

@@

\subsection{Ordinal addition}

@@

\subsubsection{Multiplication by a natural number}

@@

\section{The exponential of basis \texorpdfstring{$\omega$}{omega}}

@@

\subsection{Additive principal ordinals}

@@

\subsection{The function \texttt{phi0}}

@@

\subsection{Omega-towers and the ordinal \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\subsection{Properties of the set  \texttt{AP}}

@@

\subsubsection{Properties of the function \texorpdfstring{$\phi_0$}{phi0}}

@@

\section{Critical ordinals}

@@

\subsection{A flavor of infinity}

@@

\section{Cantor normal form}

@@

\section{An embedding of \texttt{T1} into \texttt{Ord}}

@@

\section{Related work}

@@

\chapter{The ordinal \texorpdfstring{$\Gamma_0$}{Gamma0} (first draft)}

@@

\index{Projects}
\begin{project}
Prove that a (countable)  ordinal $\alpha$ is strongly critical iff $\phi \alpha 0=\alpha$ (Theorem 13.13 of~\cite{schutte} ). 
\end{project}

@@

In the present version of this development, we will only study $\Gamma_0$ as a notation system, much more powerful than the ordinal notation for $\epsilon_0$.

@@

\section{The type \texttt{T2} of ordinal terms}

@@

\section{How big is \texorpdfstring{$\Gamma_0$}{\texttt{Gamma0}}?}

@@

\subsection{Length of a term}

@@

\section{Main functions on \texttt{T2}}

@@

Like for Cantor normal forms (see Sect.\~ref{sect:infix-plus-T1})  the definition of addition in \texttt{T2}  requires comparison between ordinal terms.

@@

\subsection{The Veblen function \texorpdfstring{$\phi$}{\texttt{phi}}}

>>>>

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, Bordeaux-INP\\with contributions by \'Evelyne Contejean,  Florian Hatat and Pascal Manoury}
\date{\today}
\title{Hydra Battles in Coq}
\makeindex
\begin{document}
\maketitle

@@

\subsection{Difference from Kirby and Paris's Work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

\subsection{Trust in our Proofs?}
\label{sect:trust-in-proofs}

@@

\subsection{Assumed Redundancy}

@@

\subsection{Main References}

@@

\subsection{State of the Development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on  notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

@@

\subsection{Typographical Conventions}

@@

\subsection{Active Links}
The  links which appear in this pdf  document lead are of three possible kinds of destination:
\begin{itemize}
\item Local links to the document itself,
\item External links, mainly to \coq's page,
\item Local links to pages generated by \texttt{coqdoc}. According to the current makefile,
  we assume that the page generated from a library \texttt{XXX/YYY.v} is stored as
the relative address \texttt{../src/html/hydras.XXX.YYY.html} (from the location of the pdf).
Thus,  active links towards our \coq{} modules may be incorrect if you got this \texttt{pdf} document otherwise than by compiling the distribution available in
\url{https://framagit.org/casteran/hydra-battles-in-coq}.

@@

\subsection{Alternative or Bad Definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

@@

\subsection{The Rules of the Game}

@@

\section{Hydras and their Representation in \emph{Coq}}
\label{sec:orgheadline48}

@@

\subsubsection{Recursive Functions on type \texttt{Hydra}}
\label{sec:orgheadline41}
\label{sec:hsize-def}

@@

\subsection{Induction Principles for Hydras}
\label{sec:orgheadline42}

@@

\subsubsection{A failed Attempt}

@@

\subsubsection{A Principle of Mutual Induction}
In order to get an appropriate induction scheme for the types 
\texttt{Hydra} and \texttt{Hydrae}, we can use  \coq{}'s  command \texttt{Scheme}.

@@

\subsubsection{A Correct Proof}

@@

\section{Relational Description of Hydra Battles}

@@

\subsection{Chopping off a Head at Distance 1 from the Foot (Relation  R1)}

@@

\subsection{Chopping off a Head at Distance \texorpdfstring{$\geq 2$}{>= 2} from the Foot (relation R2) }

@@

\subsection{Relation Associated with a Round}

@@

\subsection{Rounds and Battles}

@@

\subsection{Classes of Battles}
\label{sect:battle-classes}

@@

\subsection{Big Steps}

@@

\section{A Long Battle}
\label{sect:big-battle}

@@

\subsection{The beginning of Hostilities}
During the two first rounds, our hydra loses its two rightmost heads. Thus just before the third round, it looks like in figure~\vref{fig:hinit-plus2}.

@@

\subsection{Looking for Regularities}

@@

\subsection{Giant Steps}

@@

\subsection{A Minoration Lemma}

@@

\section{Generic Properties}

@@

\subsection{A  Small Proof of Impossibility}
\index{Maths!Proofs of impossibility}

@@

\chapter{Introduction to Ordinal Numbers and Ordinal Notations}

@@

\section{The Mathematical Point of View}

\subsection{Well-ordered Sets}
Let us start with some definitions.
A  \emph{well-ordered set} is a set provided with a binary relation $<$ which has the following properties.
\begin{description}
\item[irreflexivity] : $\forall x\in A, x\not< x$
\item[transitivity] : $\forall x\,y\,z\in A, x<y \Rightarrow y<z \Rightarrow x<z$
\item[trichotomy]: $\forall x\,y\in A, x<y \vee x = y \vee y < x$
\item[well foundedness]: $<$ is well-founded (every element of $A$ is accessible)\footnote{In classical mathematics, we would say that there is no infinite sequence $a_1>a_2> \dots a_n> a_{n+1}\dots$ in $A$. Please refer to any documenation on \coq{} for having more details on well-foundedness and accessibility.}.

@@

\subsection{Ordinal Numbers}

@@

\section{Ordinal Numbers in Coq}

@@

\section{Countable Ordinals}

@@

\section{Ordinal Notations}

@@

\subsubsection*{Ordered Types}

@@

\subsection{A Class for Ordinal Notations}

@@

\subsection{Ordinal Notations and Measures for Proving Termination}
\label{sect:measure-ON}

@@

\section{Examples of Ordinal Notations}

\subsection{The Ordinal \texorpdfstring{$\omega$}{omega}}

@@

\subsection{Sum of two Ordinal Notations}

@@

\subsection{The Ordinal \texorpdfstring{$\omega+\omega$}{omega + omega}}

@@

\subsection{Limits and Successors in \texorpdfstring{$\omega+\omega$}{omega+omega}}

@@

\section{The Ordinal \texorpdfstring{$\omega^2$}{omega^2}}

@@

\subsection{A Proof of Termination using \texorpdfstring{$\omega^2$}{omega^2}} 
\label{sect:merge-example}.

@@

\subsection{Yet Another  Proof of Impossibility}
\label{omega2-case}

@@

\subsubsection{Proof of the Inequality \texttt{m small\_h o< m big\_h}}

@@

\subsubsection{Proof of the Inequality \texttt{m big\_h o<= m small\_h} }

@@

\subsubsection{End of the Proof}

@@

\section{A Notation for Finite Ordinals}

@@

\section{Comparing two Ordinal Notations}

@@

\section{Comparing an Ordinal Notation with Schütte's Model}

@@

\section{Isomorphism of Ordinal Notations}

@@

\section{Other Ordinal Notations}

@@

\chapter[A proof of termination, using epsilon0]{A Proof of Termination, using Ordinals below \texorpdfstring{$\epsilon_0$}{Epsilon0}}

@@

\section{The Ordinal \texorpdfstring{\(\epsilon_0\)}{epsilon0}}
\label{sec:epsilon0-intro}

\subsection{Cantor Normal Form}
\index{Maths!Cantor normal form}

@@

\subsection{A Data Type for  Ordinals in Cantor Normal Form}
\label{sec:orgheadline72}
\label{sec:T1-inductive-def}

@@

\subsubsection{Finite Ordinals}
\label{sec:orgheadline67}

@@

\subsubsection{The Ordinal \(\omega\)}
\label{sec:orgheadline68}

@@

\subsubsection{The Ordinal \(\omega^\alpha\), a.k.a. \(\phi_0(\alpha)\)}
\label{sec:orgheadline70}

@@

\subsubsection{The Hierarchy of \(\omega\)-towers:}
\label{sec:orgheadline71}

@@

\subsection{Comparison between Ordinal Terms}
\label{sec:orgheadline73}

@@

\subsubsection{A Predicate for Characterizing Normal Forms}
\label{sect:t1-nf}

@@

\subsection{Making Normality Implicit}
  We would like to get rid of terms of type \texttt{T1} which are not in Cantor normal form.
A simple way to do this is to consider statements of the form 
\texttt{forall alpha: T1, nf alpha -> $P$ alpha}, where $P$ is a predicate over type \texttt{T1}, like in the following lemma \footnote{Ordinal addition is formally defined a little later (page~\ref{sect:infix-plus-T1})}.

@@

\subsubsection{A Sigma Type for \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\subsection{Syntactic Definition of Limit and Successor Ordinals}

@@

\subsubsection{Addition and Multiplication}

@@

\subsection{Pretty Printing Ordinals in Cantor Normal Form}
\label{sect:ppT1}

@@

The following data type defines an abstract syntax for more readable ordinals terms in Cantor normal form:

@@

\subsubsection{Arithmetic on Type \texttt{E0}}

@@

\section{Well-foundedness and Transfinite Induction}

@@

\subsection{About  Well-foundedness}
\label{sec:orgheadline82}
   In order to use \texttt{T1} for proving termination results,
we need to prove that  our order \texttt{<} is well-founded. Then we will get \emph{transfinite induction} for free.

@@

\subsubsection{A First Attempt}
\label{sec:orgheadline77}
\index{Coq!Techniques!Well-founded induction}

@@

\subsubsection{Using a Stronger Inductive Predicate.}
\label{sec:orgheadline78}
  Instead of trying to prove directly that any ordinal term \(\alpha\) in normal form is accessible
through \texttt{LT}, we propose to show first that any well formed 
term of the form \(\omega^\alpha\times(n+1)+\beta\) is accessible (which is a stronger result).

@@

\subsection{An Ordinal Notation for  \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\section{A Variant for Hydra Battles}

\subsection{Natural Sum (a.k.a. Hessenberg's  Sum)}
\label{sec:orgheadline87}
\label{hydra-variant}

@@

\subsubsection{Rewriting Lemmas}
\label{sec:orgheadline86}

@@

\index{Projects}
\begin{project}
Compare \texttt{oplus}'s definition (with inner fixpoint) with other possibilities
(\texttt{coq-equations}, \texttt{Function}, etc.).
\end{project}
\subsection{More Theorems on Hessenberg's Sum}

@@

\subsubsection{Commutativity, Associativity}

@@

\subsection{A Measure for Hydra-battle Termination}

@@

\chapter[The Ketonen-Solovay machinery]{Strolling inside \texorpdfstring{$\epsilon_0$}{Epsilon0}: The Ketonen-Solovay Machinery\label{ks-chapter}}
\label{chap:ketonen}
\index{Maths!Ketonen-Solovay machinery}

@@

\subsubsection{Mathematical Definition of Canonical Sequences}

@@

\subsubsection{Canonical Sequences in Coq}

@@

\subsection{Basic Properties of Canonical Sequences}

@@

\subsubsection{Canonical Sequences and the Order $<$}

@@

\subsubsection{Limit Ordinals are Really Limits}
The following theorem states that any limit ordinal $\lambda<\epsilon_0$ 
is the limit of the sequence \showmath{\canonseq{\lambda}{i}\;(1\le i)}.

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : Paths}
\index{Maths!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\subsection{Formal Definition}

@@

\subsection{Existence of a Path}

@@

\subsection{Paths and Hydra Battles}
\label{KS-o2h}

@@

\section{A  Proof of Impossibility}

@@

\subsection{The case of Free Battles}
\label{sec:free-battles-case}
Let us now consider that $B$ is instantiated to \texttt{free} (which means that we are considering proofs of termination of \emph{all} battles). The following lemmas are proved in Module~\href{../src/html/hydras.Hydra.Epsilon0_Needed_Free.html}{Hydra.Epsilon0\_Needed\_Free}.
The case $B=\texttt{standard}$ is studied in section~\vref{std-case}.

@@

\section{The Case of Standard Battles}
\label{sec:standard-intro}\label{std-case}
One may wonder if our theorem holds also in the framework of standard battles. Unfortunately, its proof relies on the lemma \texttt{LT\_to\_round\_plus} of
Module~\href{../src/html/hydras.Hydra.O2H.html}{Hydra.O2H}.

@@

\subsection{Paths with a Constant Index}

@@

\subsubsection{Sketch of Proof of \texttt{Cor12}}
\index{Maths!Transfinite induction}

@@

\subsection{Casting Paths with a Constant Index into a Standard Path}

@@

\subsubsection{Sketch of Proof of \texttt{constant\_to\_standard\_path}}

@@

\subsection{Back to Hydras}
\label{sec:standard-battles-cases}
We are now able to complete our proof that there exists no bounded variant for proving the termination of standard hydra battles. This proof can
be consulted in the module 
\url{../src/html/hydras.Hydra.Epsilon0_Needed_Std.html}.
Please note that it has the same global structure as in section\ref{sec:free-battles-case} 
% ICI !
Applying the  lemmas  \texttt{Lemma2\_6\_1} of the module 
\href{../src/html/hydras.Epsilon0.Paths.html\#Lemma2_6_1}%
{\texttt{Lemma2\_6\_1}}   and 
\href{../src/html/hydras.Epsilon0.Paths.html\#constant_to_standard_path}%
{\texttt{constant\_to\_standard\_path}},
we can convert any inequality $\beta<\alpha<\epsilon_0$ into a standard path from
$\alpha$ to  $\beta$, then into a fragment of a standard battle from 
$\iota(\alpha)$ to $\iota(\beta)$.

@@

%---------------------------------------------------------------------
\chapter{Large Sets and Rapidly Growing Functions}\label{chap:alpha-large}

@@

\section{The Length of Minimal Large Sequences}

@@

\subsection{Formal Specification}

@@

\subsection{Abstract Properties}

@@

\subsubsection{The  Ordinal Zero}
\label{sect:L-spec-zero}
The base case is directly a consequence of the specification.

@@

\subsubsection{Successor Ordinals}
\label{sect:L-spec-succ}
Let $\beta$ be some ordinal, and assume the arithmetic function $f$ satisfies 
the specification $(\texttt{L\_spec}\;\beta)$.  Let $k$ be any natural number.
Any path from $\texttt{succ}\,\beta$ to $0$ starting at $k+1$ can be decomposed into a first step from $\texttt{succ}\,\beta$ to $\beta$, then a path from
$\beta$ at $k+2$ to $0$. 
By hypothesis the interval $[k+2, f(k+2)-1]$ is $\beta$-mlarge.
But the interval $[k+1, f(k+2)-1]$ is the concatenation of the singleton
$\{k+1\}$ and the interval $[k+2, f(k+2)-1]$.
So, the function $\lambda\,k.\,f(k+1)$ satisfies the specification $\texttt{L\_spec}\,\beta$.

@@

\subsubsection{Limit Ordinals}
\label{sect:L-spec-lim}

@@

\subsection{First Results}

@@

\subsubsection{Finite Ordinals}

@@

\subsubsection{The First Limit Ordinal  \texorpdfstring{$\omega$}{omega}}

@@

\subsubsection{Going Further}
Let us consider a last example, ``computing'' $L_{\omega^3}$.
Since the canonical sequence associated with this ordinal is composed of the
$\omega^2\times i\;(i\in\mathbb{N}_1)$, we have to study this sequence.

@@

\subsubsection{Back to Hydra Battles}

@@

\section{The Wainer-Hardy Hierarchy (Functions \texorpdfstring{$H_\alpha$}{\texttt{H\_alpha}})}

@@

\subsection{Hardy Functions in \texttt{Coq}}

@@

\subsection{First  Steps of the Hardy hierarchy}
Using rewrite rules from \texttt{H\_eq1} to \texttt{H\_eq4}, we can explore the functions $H_\alpha$ for some small values of $\alpha$.

\subsubsection{Finite Ordinals}

@@

\subsubsection{New Limits}

@@

\subsubsection{A Numerical Example}

@@

\subsection{Abstract Properties of H-functions}
~\label{sect:H-alpha-prop}

@@

\subsubsection{Back to Hydras}

@@

\section{The Wainer Hierarchy (Functions \texorpdfstring{$F_\alpha$}{F\_alpha})}
\label{sect:wainer}

@@

%----------------------------------------------------------------------
\chapter[Countable Ordinals (after Sch\"{u}tte)]{Kurt Schütte's Axiomatic Definition of Countable Ordinals}

@@

\section{Additional  Axioms}

@@

\subsubsection{Classical Logic}

@@

\subsubsection{Description Operators}

@@

\subsubsection{The Definition of zero}

@@

\section{The  Successor Function}

@@

\section{Finite Ordinals}

@@

\section{The Definition of \texttt{omega}}
In order to define $\omega$, the first infinite ordinal, we use an operator which
``returns'' the least upper bound (if it exists) of a subset $X\subseteq \mathbb{O}$.
For that purpose, we first use a predicate:
(\texttt{is\_lub $D$ \textit{lt} $X$ $a$}) if $a$ belongs to $D$ and is the least 
upper bound  of $X$ (with respect to \textit{lt}).

@@

\subsection{Ordering Functions and Ordinal Addition}

@@

\subsection{Ordinal Addition}

@@

\subsubsection{Multiplication by a Natural Number}

@@

\section{The Exponential of Basis \texorpdfstring{$\omega$}{omega}}

@@

\subsection{Additive Principal Ordinals}

@@

\subsection{The Function \texttt{phi0}}

@@

\subsection{Omega-towers and the Ordinal \texorpdfstring{$\epsilon_0$}{epsilon0}}

@@

\subsection{Properties of the Set  \texttt{AP}}

@@

\subsubsection{Properties of the Function \texorpdfstring{$\phi_0$}{phi0}}

@@

\section{Critical Ordinals}

@@

\subsection{A flavor of Infinity}

@@

\section{Cantor Normal Form}

@@

\section{An Embedding of \texttt{T1} into \texttt{Ord}}

@@

\section{Related Work}

@@

\chapter{The Ordinal \texorpdfstring{$\Gamma_0$}{Gamma0} (first draft)}

@@

\index{Projects}
\begin{project}
Prove that a (countable)  ordinal $\alpha$ is strongly critical iff 
$\phi_\alpha(0)=\alpha$ (Theorem 13.13 of~\cite{schutte} ). 
\end{project}

@@

In the present version of this development, we  only study $\Gamma_0$ as a notation system, much more powerful than the ordinal notation for $\epsilon_0$.

@@

\section{The Type \texttt{T2} of Ordinal Terms}

@@

\section{How Big is \texorpdfstring{$\Gamma_0$}{\texttt{Gamma0}}?}

@@

\subsection{Length of a Term}

@@

\section{Main Functions on \texttt{T2}}

@@

Like for Cantor normal forms (see Sect.~\ref{sect:infix-plus-T1}),  the definition of addition in \texttt{T2}  requires comparison between ordinal terms.

@@

\subsection{The Veblen Function \texorpdfstring{$\phi$}{\texttt{phi}}}
